---
title: "Pollinator-dependence analytics for exclusion experiments"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Pollinator-dependence analytics for exclusion experiments}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
set.seed(1)
library(pollinatoR)
```

## What this package is for

`pollinatoR` estimates the contribution of insect pollinators to crop yield from
*pollinator-exclusion experiments* -- paired open and caged (exclusion) plots, and
related contrasts -- and relates that contribution to the pollinator community. The
methods are general to open-versus-exclusion designs; canola is one application, not
a constraint.

The package is a small set of verbs around one declared study object. This vignette
runs every one of them end to end on a simulated canola-shaped dataset, so each step
produces real output you can read.

| Verb | Question it answers |
| --- | --- |
| `pollin_spec()` | What is the study -- outcomes, treatment, nesting, exposures? |
| `pollin_contribution()` | How much yield does pollinator access add, with uncertainty? |
| `pollin_fdr()` | Which contributions survive multiple-testing control? |
| `pollin_moderation()` | Which insects drive the contribution, free of unit confounders? |
| `pollin_rii()` / `pollin_index()` | The dependence index, with an interval, not a bare ratio. |
| `pollin_guard_spurious()` | Is an index-versus-yield slope real or mechanical coupling? |
| `pollin_within_between()` | The Mundlak split that makes moderation confounder-robust. |
| `pollin_compensation()` | Do yield components trade off against each other? |
| `pollin_validate()` | Does the data pass structural integrity checks? |

## A simulated study

One open-versus-cage design: 10 sites, 16 plants each, two insect exposures, and two
yield-component outcomes. Open access lifts pod set, and does so more where honeybees
are abundant -- the moderation signal we will recover.

```{r}
ns <- 10L
per <- 16L
site <- rep(sprintf("s%02d", seq_len(ns)), each = per)
trt <- rep(rep(c("Cage", "Open"), each = per / 2L), ns)
honeybee <- rep(rpois(ns, 8), each = per) + rpois(ns * per, 1)
wild <- rep(rpois(ns, 4), each = per) + rpois(ns * per, 1)
site_eff <- rnorm(ns, 0, 0.3)[as.integer(factor(site))]
open <- trt == "Open"
pod_mu <- exp(1.8 + 0.35 * open + 0.015 * honeybee * open + site_eff)
pods <- rpois(ns * per, pod_mu)
seeds_per_pod <- rpois(ns * per, exp(2.6 + 0.12 * open + site_eff))
canola <- data.frame(site, trt, honeybee, wild, pods, seeds_per_pod)
head(canola)
```

## Declare the study once

```{r}
spec <- pollin_spec(
  data = canola,
  outcomes = c("pods", "seeds_per_pod"),
  treatment = "trt",
  control = "Cage",
  random = "site",
  exposures = c("honeybee", "wild"),
  latent_unit = "site" # carries the random treatment slope
)
spec
```

## The contribution, with uncertainty

`pollin_contribution()` fits a `glmmTMB` mixed model per outcome (family chosen
heuristically) and reports the open-minus-exclusion contrast through
`marginaleffects`, with a confidence interval and the standard deviation of the
random treatment slope -- a reportable measure of how much the contribution varies
between units.

```{r, warning = FALSE, message = FALSE}
contribution <- pollin_contribution(spec)
contribution
```

With more than one outcome, control the false-discovery rate across them:

```{r}
contribution$table$fdr <- pollin_fdr(contribution$table$p.value)
contribution$table[, c("outcome", "estimate", "p.value", "fdr")]
```

## Which insects drive it

`pollin_moderation()` enters each exposure through a within-between (Mundlak) split.
The within-unit treatment-by-exposure term is the confounder-robust estimate: it is
identified from variation inside a site, so an unrecorded site-level genotype cannot
masquerade as a pollinator effect.

```{r, warning = FALSE, message = FALSE}
moderation <- pollin_moderation(spec, outcome = "pods")
moderation
```

The split itself is available directly; `between` is the unit mean and `within` the
deviation within a unit, robust to any unit-level confounder.

```{r}
wb <- pollin_within_between(canola$honeybee, group = canola$site)
str(wb)
```

## Uncertainty-aware indices

Per-site open and exclusion means give a dependence index with an interval, rather
than a single ratio. `pollin_rii()` returns the Relative Interaction Index with a
bootstrap interval; `pollin_index()` adds the log response ratio with a closed-form
interval.

```{r}
open_mean <- tapply(canola$pods[open], canola$site[open], mean)
cage_mean <- tapply(canola$pods[!open], canola$site[!open], mean)
pollin_rii(open_mean, cage_mean, n_boot = 500, seed = 1)
pollin_index(open_mean, cage_mean, n_boot = 500, seed = 1)
```

## Guard against spurious index-versus-yield slopes

Regressing a dependence index on absolute (open-arm) yield is hazardous: the open
term sits on both axes, so a positive slope is partly mechanical (Pearson 1897). The
guard permutes the exclusion arm while holding the open arm fixed and reports how much
of the slope is coupling.

```{r}
pollin_guard_spurious(open_mean, cage_mean, n_perm = 200, seed = 1)
```

A permutation p-value near 1 means the observed slope is indistinguishable from the
coupling null -- spurious. Supply `decouple =` (the exclusion arm, or an independent
vigour covariate) to regress the index on an axis that shares no term.

## Yield-component compensation

When pollinators lift one yield component, plants may down-regulate another.
`pollin_compensation()` summarises the trade-off structure across components as a
correlation matrix.

```{r}
pollin_compensation(canola[, c("pods", "seeds_per_pod")])
```

## Integrity checks

`pollin_validate()` runs structural checks -- missingness, expected factor levels, and
user-supplied identities -- before any model is fitted.

```{r}
pollin_validate(spec, expect_levels = list(trt = c("Cage", "Open")))
```
