---
title: "A walkthrough of `pollinatoR`"
subtitle: "Pollinator-dependence analytics for exclusion experiments, end to end"
author: "Max Moldovan"
date: "2026-06-05"
output:
  pdf_document:
    latex_engine: xelatex
    toc: true
    toc_depth: 2
    number_sections: true
    highlight: tango
geometry: margin=2.6cm
header-includes:
  - \usepackage{lmodern}
  - \usepackage{microtype}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  fig.align = "center"
)
set.seed(1)
library(pollinatoR)
```

# What this walkthrough is

`pollinatoR` turns a pollinator-exclusion experiment -- paired open and caged plots --
into a defensible estimate of what insect pollinators contribute to yield, with the
uncertainty carried through rather than discarded. This document runs the whole package
on one worked dataset, from declaring the study to the integrity checks a reviewer would
demand. A reader should be able to follow it in ten minutes and reproduce it in thirty.

Two voices run through it. The *analyst* wants a number: how much does pollinator
access add, and which insects drive it. The *reviewer* distrusts that number and looks
for the ways a pollination study usually goes wrong -- a ratio index regressed on a yield
axis it shares a term with, an effect that is really an unrecorded site difference, a
yield component that was traded away rather than gained.

The data here are simulated, not the field trial: every value below comes from the
`set.seed(1)` block in the next section, so nothing in this walkthrough touches real or
restricted records.

# The worked dataset

A modest open-versus-cage design: ten sites, sixteen plants per site, two insect
exposures (honeybee and wild-insect counts), and two yield components (pod number and
seeds per pod). The data-generating process is deliberate. Open access lifts pod set,
and it lifts pods more where honeybees are abundant -- a treatment-by-exposure
interaction the analyst should recover and the reviewer should be able to trust.

```{r data}
ns <- 10L
per <- 16L
Site <- rep(sprintf("s%02d", seq_len(ns)), each = per)
trt <- rep(rep(c("Cage", "Open"), each = per / 2L), ns)
honeybee <- rep(rpois(ns, 8), each = per) + rpois(ns * per, 1)
wild <- rep(rpois(ns, 4), each = per) + rpois(ns * per, 1)
site_eff <- rnorm(ns, 0, 0.3)[as.integer(factor(Site))]
open <- trt == "Open"
pods <- rpois(ns * per, exp(1.8 + 0.35 * open + 0.015 * honeybee * open + site_eff))
seeds_per_pod <- rpois(ns * per, exp(2.6 + 0.12 * open + site_eff))
canola <- data.frame(Site, trt, honeybee, wild, pods, seeds_per_pod)
head(canola)
```

# Act I -- the analyst

## Declare the study once

Every later call reads from one `pollin_spec()` object, so the study definition lives in
exactly one place. The `latent_unit` is the unit that carries a random treatment slope:
the site-to-site spread in the contribution becomes a parameter rather than noise.

```{r spec}
spec <- pollin_spec(
  data = canola,
  outcomes = c("pods", "seeds_per_pod"),
  treatment = "trt",
  control = "Cage",
  random = "Site",
  exposures = c("honeybee", "wild"),
  latent_unit = "Site"
)
spec
```

## Estimate the contribution

`pollin_contribution()` fits a `glmmTMB` mixed model for each outcome, choosing a count,
proportion, or continuous family from the data, and reports the open-minus-exclusion
contrast through `marginaleffects`. The estimate is a contribution on the response
scale, with a confidence interval and `slope_sd` -- the standard deviation of the random
treatment slope, which says how much the contribution itself varies between sites.

```{r contribution}
contribution <- pollin_contribution(spec)
contribution
```

With two outcomes the analyst is already testing more than one hypothesis, so the
p-values are brought under false-discovery-rate control before anything is called
significant.

```{r fdr}
contribution$table$fdr <- pollin_fdr(contribution$table$p.value)
contribution$table[, c("outcome", "estimate", "p.value", "fdr")]
```

## Attribute it to the community

The next question -- which insects drive the contribution -- is where naive analyses
fail, because a site with many honeybees may differ in a dozen unrecorded ways.
`pollin_moderation()` answers it with a within-between (Mundlak) decomposition: each
exposure is split into its site mean and the deviation within a site, and only the
within-site treatment-by-exposure term is read as the pollinator effect.

```{r moderation}
moderation <- pollin_moderation(spec, outcome = "pods")
moderation
```

The split is available on its own through `pollin_within_between()`. The `between`
component is the site mean; the `within` component is the part identified inside a site,
robust to any site-level confounder.

```{r within-between}
wb <- pollin_within_between(canola$honeybee, group = canola$Site)
str(wb)
```

# Act II -- the reviewer

## Indices, but with an interval

The reviewer accepts the dependence index as a summary, but not as a bare ratio.
`pollin_rii()` returns the Relative Interaction Index with a bootstrap interval, and
`pollin_index()` adds the log response ratio with a closed-form interval. Here the index
is built from per-site open and exclusion pod means.

```{r indices}
open_mean <- tapply(canola$pods[open], canola$Site[open], mean)
cage_mean <- tapply(canola$pods[!open], canola$Site[!open], mean)
pollin_rii(open_mean, cage_mean, n_boot = 500, seed = 1)
pollin_index(open_mean, cage_mean, n_boot = 500, seed = 1)
```

## The coupling trap

The reviewer's sharpest objection: a paper that regresses a dependence index on absolute
yield is partly measuring an identity, because the open-arm term sits on both axes
(Pearson 1897). `pollin_guard_spurious()` permutes the exclusion arm while holding the
open arm fixed, so the permutation distribution is exactly the mechanical-coupling null.

```{r guard}
pollin_guard_spurious(open_mean, cage_mean, n_perm = 200, seed = 1)
```

A permutation p-value near one means the observed slope is indistinguishable from
coupling -- it should not be reported as a finding. The `decouple =` argument lets the
analyst regress the index on an axis that shares no term (the exclusion arm, or an
independent vigour covariate) and recover an honest slope.

## Was the yield gained or merely moved

A contribution to one component can be a true gain or a re-allocation from another.
`pollin_compensation()` summarises the trade-off structure among yield components as a
correlation matrix, so a negative off-diagonal flags compensation rather than gain.

```{r compensation}
pollin_compensation(canola[, c("pods", "seeds_per_pod")])
```

## Integrity before inference

Finally, the checks that should run before any model is fitted: missingness, the
factor levels the design promises, and any identity the analyst asserts about the data.

```{r validate}
pollin_validate(spec, expect_levels = list(trt = c("Cage", "Open")))
```

# Reproduce it yourself

Everything above runs from a clean session in well under a minute:

```r
# install.packages("pak")
pak::pak("max578/pollinatoR")
library(pollinatoR)

# then paste the "worked dataset" block and run each section in order.
```

The only stochastic steps are the data simulation and the bootstrap and permutation
intervals; each carries an explicit seed, so the numbers reproduce exactly.

# Notes and limitations

The dataset is synthetic and small, chosen so the walkthrough runs quickly and the
recovered signal can be checked against the known data-generating process. On real data
the family heuristic should be confirmed against residual diagnostics, the random
treatment slope needs enough units to be estimable, and the within-between split assumes
the exposures are measured without large error. The package estimates a contribution
under the design it is given; it does not repair a design that lacks a credible
exclusion contrast, and the spurious-regression guard flags a coupled regression but
cannot rescue a question that should not have been asked as a ratio in the first place.

# Session information

```{r session}
sessionInfo()
```
